---
title: "Class 11 Lab Session"
author: "Nicholas Yousefi"
format: pdf
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble <https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39863622-39954623;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel>

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Lets now dig into this further.

# Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

How many samples do we have? (**read file into R**)
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

**Sample size for each genotype:**

```{r}
table(expr$geno)
```

**Median Expression Level for each genotype:**

```{r}
findMedianExp <- function(genotype) {
  expression <- expr$exp[expr$geno == genotype]
  return(median(expression))
}
findMedianExp("A/A")
findMedianExp("A/G")
findMedianExp("G/G")
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

Lets make a boxplot.

```{r}
ggplot(expr) +
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=T)
```

It looks like people with the G/G genotype express ORMDL3 less than people with the A/A genotype. This SNP does appear to affect the expression of ORMDL3. 